---
title: "Normalizing and denoising protein expression data from droplet-based single-cell profiling"
output: github_document
---

### Analysis code to reproduce manuscript results  
Mulè MP, Martins AJ, Tsang JS. Normalizing and denoising protein expression data from droplet-based single cell profiling. bioRxiv. 2020;2020.02.24.963603.  
repository author: Matt Mulè mulemp@nih.gov    
general contact: John Tsang john.tsang@nih.gov    

### The dsb R package 
Analysis in this paper supports the dsb R package for CITE-seq data denoising and normalization
[Download the dsb package from the NIAID github](https://github.com/niaid/dsb)

### to install the dsb package 
install dsb directly from github: 
```{r, eval = FALSE}
require(devtools); devtools::install_github(repo = 'niaid/dsb')
library(dsb)
```


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
### Introduction

This analysis code pipeline is for reproducing the analysis in the paper: **Normalizing and denoising protein expression data from droplet based single cell profiling** The paper introduces the experiments and statistical modeling underlying the dsb normalization method for CITE-seq protein data. The dsb R package (see above) is a lightweight R package and is the first dedicated method for normalizing and denoising antibody based protein expression data from droplet based single cell experiments (CITE-seq, REAP-seq, Mission Bio Tapestri etc). The method was developed in [John Tsang’s Lab](https://www.niaid.nih.gov/research/john-tsang-phd) by Matt Mulè, Andrew Martins and John Tsang.  

This analysis workflow presented in the preprint can be run on a laptop with 16GB RAM. Note in this reproducible workflow, the directory containing the dsb_normalization.rproj file acts like the root directory by using the 'here' R package. To reproduce analysis results, simply download the github repository, and add data described below into a folder called data/V2_Data from any folder from which you run the analysis. Each R script is a self-contained analysis script reading the data from the /data folder and writing figures or results files within a generated_data directory within each analyis subdirectory as outlined below.

### Software package versions used in this analysis 

Please see full session info at the end of this script 

R version 3.5.3 
mclust_5.4.5    
reticulate_1.12  
umap_0.2.3.1    
magrittr_1.5    
forcats_0.4.0   
stringr_1.4.0   
dplyr_0.8.5    
purrr_0.3.3     
readr_1.3.1     
tidyr_1.0.2     
tibble_2.1.1    
tidyverse_1.2.1  
Seurat_2.3.4    
Matrix_1.2-15  
cowplot_0.9.4   
ggplot2_3.1.1   
here_0.1       

### R packages used in this this analysis  
```{r, eval=FALSE}
# modeling  / analysis / helper functions  
library(Seurat)
library(tidyverse)
library(magrittr)
library(here)
library(mclust)
library(dsb) # see above for installation instructions.

# visualization
library(ggrepel)
library(ggridges)
library(pals)

# For umap only, must have python virtual env for this particular pipeline, for example run: 
# python must be installed 
virtualenv_create("r-reticulate")
virtualenv_install("r-reticulate", "umap-learn")
use_virtualenv("r-reticulate")
library(umap)
library(reticulate)
library(umap)
```

The data directory: *data/V2_Data/* contains the starting raw data needed for analysis. Prior to running analysis confirm the data are in the correct repository. 

```{r}
suppressMessages(library(here))
# all STARTING data 
data_ = c(
  "CITEseq_raw_PMID32094927_seurat2.4.rds", 
  "unstained_control_singlets.rds"
)
stopifnot(data_ %in% list.files(here("data/V2_Data/")))

# background drop data 
data_ = c(
"adt_neg_dmx_list.rds",
"adt_neg_dmx.rds",
"adt_neg_full_list.rds",
"adt_neg_full_qc.rds"
)
stopifnot(data_ %in% list.files(here("data/V2_Data/background_data/")))

# pipeline scripts scripts 
scripts_ = c(
  "protein_annotation_functions.r",
  "geneplot4.R"
)
stopifnot(scripts_ %in% list.files(here("V2/functions/")))


```


### dsb normalize PBMC data from 20 individuals.  
V2/dsb_normalize_cluster_pipeline/

```{r, eval=FALSE}
source(here("V2/dsb_normalize_cluster_pipeline/1_dsb_normalize.r"))
# creates data drectory here("V2/dsb_normalize_cluster_pipeline/generated_data/") containing results 
# "dsb2_normlog10_list_bybatch.rds"
# "h1_d0_singlets_ann_Seurat2.4_dsbnorm.rds"
```

### run umap  
reads data from:  
V2/dsb_normalize_cluster_pipeline/generated_data/h1_d0_singlets_ann_Seurat2.4_dsbnorm.rds
```{r, eval=FALSE}
source(here("V2/dsb_normalize_cluster_pipeline/2_run_umap.r"))
```

### Figure generation   

```{r, eval=FALSE}
source(here("V2/dsb_normalize_cluster_pipeline/3_figure_generation.r"))
source(here("V2/dsb_normalize_cluster_pipeline/4_manual_gate_plots.r"))
```

### dsb process  
This section illustrates the dsb process in steps and assesses underlying modeling assumptions 

```{r, eval=FALSE}
source(here("V2/dsb_process_plots/6_mean_isotype_v_mean_control.R"))
source(here("V2/dsb_process_plots/6a_isotype_figure_generation.r"))
source(here("V2/dsb_process_plots/7_neg_control_plots.R"))
source(here("V2/dsb_process_plots/8_mixture_fits.r"))

```


### batch norm analysis 
Assessment of single vs multi batch normalization and sensitivity of each normalization scheme to defining background with hashing or library size distributions. 
```{r, eval=FALSE}
source(here("V2/parameter_sensitivity/empty_drop_threshold_batch.r"))
```

### 10X data analysis 

Read Cell Ranger raw output, select negative drops, run DSB normalization, run each normalization modeling step separately, cluster cells and plot distributions across clusters and on biaxial gates. Test underlying modeling assumptions for each external dataset. 

```{r, eval=FALSE}

# 10K v3 data 
source(here("V2/10x_analysis/10x_pbmc_10k_V3.r"))
source(here("V2/10x_analysis/10x_pbmc_10k_V3_figure_generation.r"))
# 5k V3 data 
source(here("V2/10x_analysis/10x_pbmc_5k_V3.r"))
source(here("V2/10x_analysis/10x_pbmc_5k_V3_figure_generation.r"))
# Next Gem data 
source(here("V2/10x_analysis/10x_pbmc_NextGem.r"))
source(here("V2/10x_analysis/10x_pbmc_NextGem_figure_generation.r"))
# 5 prime data 
source(here("V2/10x_analysis/10x_pbmc_5prime_5k.r"))
source(here("V2/10x_analysis/10x_pbmc_5prime_5k_figure_generation.r"))

```


# Sessioninfo
R version 3.5.3 

attached base packages: stats graphics grDevices utils datasets methods base

other attached packages: mclust_5.4.5 reticulate_1.12 umap_0.2.3.1 magrittr_1.5 forcats_0.4.0 stringr_1.4.0 dplyr_0.8.5
purrr_0.3.3 readr_1.3.1 tidyr_1.0.2 tibble_2.1.1 tidyverse_1.2.1 Seurat_2.3.4 Matrix_1.2-15
cowplot_0.9.4 ggplot2_3.1.1 here_0.1

loaded via a namespace (and not attached): readxl_1.3.1 snow_0.4-3 backports_1.1.4 Hmisc_4.2-0 plyr_1.8.4
igraph_1.2.4.1 lazyeval_0.2.2 splines_3.5.3 inline_0.3.15 digest_0.6.25
foreach_1.4.4 htmltools_0.3.6 lars_1.2 rsconnect_0.8.16 gdata_2.18.0
checkmate_1.9.3 cluster_2.0.7-1 mixtools_1.1.0 ROCR_1.0-7 modelr_0.1.4
matrixStats_0.54.0 R.utils_2.8.0 askpass_1.1 prettyunits_1.0.2 colorspace_1.4-1
rvest_0.3.4 haven_2.1.0 xfun_0.7 callr_3.2.0 crayon_1.3.4
jsonlite_1.6 survival_2.43-3 zoo_1.8-6 iterators_1.0.10 ape_5.3
glue_1.3.1 gtable_0.3.0 pkgbuild_1.0.3 kernlab_0.9-27 rstan_2.19.3
prabclus_2.3-1 DEoptimR_1.0-8 scales_1.0.0 mvtnorm_1.0-10 bibtex_0.4.2
Rcpp_1.0.1 metap_1.1 dtw_1.20-1 htmlTable_1.13.1 foreign_0.8-71
bit_1.1-14 proxy_0.4-23 SDMTools_1.1-221.1 Formula_1.2-3 stats4_3.5.3
tsne_0.1-3 StanHeaders_2.21.0-1 htmlwidgets_1.3 httr_1.4.0 gplots_3.0.1.1
RColorBrewer_1.1-2 fpc_2.2-1 acepack_1.4.1 modeltools_0.2-22 ica_1.0-2
pkgconfig_2.0.2 loo_2.3.1 R.methodsS3_1.7.1 flexmix_2.3-15 nnet_7.3-12
tidyselect_0.2.5 rlang_0.4.5 reshape2_1.4.3 cellranger_1.1.0 munsell_0.5.0
tools_3.5.3 cli_1.1.0 generics_0.0.2 broom_0.5.2 ggridges_0.5.1
evaluate_0.14 yaml_2.2.0 npsurv_0.4-0 processx_3.3.1 knitr_1.23
bit64_0.9-7 fitdistrplus_1.0-14 robustbase_0.93-5 caTools_1.17.1.2 RANN_2.6.1
packrat_0.5.0 pbapply_1.4-0 nlme_3.1-137 R.oo_1.22.0 xml2_1.2.0
hdf5r_1.2.0 compiler_3.5.3 rstudioapi_0.10 png_0.1-7 lsei_1.2-0
stringi_1.4.3 ps_1.3.0 lattice_0.20-38 vctrs_0.2.4 pillar_1.4.1
lifecycle_0.1.0 Rdpack_0.11-0 lmtest_0.9-37 data.table_1.12.2 bitops_1.0-6
irlba_2.3.3 gbRd_0.4-11 R6_2.4.0 latticeExtra_0.6-28 KernSmooth_2.23-15
gridExtra_2.3 codetools_0.2-16 MASS_7.3-51.1 gtools_3.8.1 assertthat_0.2.1
openssl_1.4 rprojroot_1.3-2 withr_2.1.2 diptest_0.75-7 parallel_3.5.3
doSNOW_1.0.16 hms_0.4.2 grid_3.5.3 rpart_4.1-13 class_7.3-15
rmarkdown_1.13 segmented_0.5-4.0 Rtsne_0.15 lubridate_1.7.4 base64enc_0.1-3


### data initialization script. 
The code below was run outside of this workflow for convenience, it is shown here for reference. This step reformats the dataset hosted at the [figshare repository](https://nih.figshare.com/collections/Data_and_software_code_repository_for_Broad_immune_activation_underlies_shared_set_point_signatures_for_vaccine_responsiveness_in_healthy_individuals_and_disease_activity_in_patients_with_lupus_Kotliarov_Y_Sparks_R_et_al_Nat_Med_DOI_https_d/4753772) that is associated with the [manuscript](https://doi.org/10.1038/s41591-020-0769-8) where we initially reported this dataset. The script below removes the protein normalized data slot which used an earlier version of the DSB packge for normaliation. In addition normaized RNA data, metadata, clustering graphs tSNE results are removed to reduce object size and, cell type annotations as reported in our (link above) are added to the object as metadata. 

```{r, eval=FALSE}
suppressMessages(library(Seurat)) 
suppressMessages(library(tidyverse))
suppressMessages(library(magrittr))
suppressMessages(library(here))

datapath = here("data/V2_Data/")

# clear unneeded data slots tsne cell embeddings, unneeded metadata 
h1 = readRDS(file = "data/V2_Data/H1_day0_scranNorm_adtbatchNorm_dist_clustered_TSNE_labels.rds")
h1@dr$protpca = NULL
h1@dr$tsne_p10 = NULL
h1@dr$tsne_p50 = NULL
h1@dr$tsne_p90 = NULL
h1@dr$tsne_p130 = NULL
h1@dr$tsne_p170 = NULL

h1@assay$HTO = NULL
h1@assay$CITE@data = NULL
h1@assay$CITE@scale.data = NULL
h1@data = NULL
# remove unused cells RNA data from raw slot
h1 = h1 %>% SubsetData(subset.raw = TRUE)

# remove unused additional metadata 
mds = read_delim(file = here("init_data_ignore/md_strip.txt"),delim = "\t")
vrm = mds$vars_
for (i in 1:length(vrm)) {
  vr = vrm[i]
  h1@meta.data[[vr]] = NULL
}

# add cluster labels from Fig 6 of https://www.nature.com/articles/s41591-020-0769-8 for reference in dsb paper.
cmd = read_delim(file = paste0(datapath, "clustree_node_labels_withCellTypeLabels.txt"), delim = '\t')
cmd1 = cmd %>% filter(clustering == "p3_dist_1")
cmd3 = cmd %>% filter(clustering == "p3_dist_3")

mdn = h1@meta.data %>% 
  mutate(celltype_label_1 = plyr::mapvalues(x = h1@meta.data$p3_dist_1, from = cmd1$cluster, to =cmd1$`Cell Type label`)) %>% 
  mutate(celltype_label_3 = plyr::mapvalues(x = h1@meta.data$p3_dist_3, from = cmd3$cluster, to =cmd3$`Cell Type label`)) %>% 
  select(barcode_check, celltype_label_3, celltype_label_1) %>% 
  column_to_rownames("barcode_check")

h1 = h1 %>% AddMetaData(metadata = mdn)

# save this starting object for DSB paper analysis  
saveRDS(h1, file = paste0(datapath, "CITEseq_raw_PMID32094927_seurat2.4.rds"))
```

### Project directory structure

├── README.Rmd  
├── README.md  
├── V2  
│   ├── 10x_analysis  
│   │   ├── 10x_pbmc_10k_V3.r  
│   │   ├── 10x_pbmc_10k_V3_figure_generation.r  
│   │   ├── 10x_pbmc_5k_V3.r  
│   │   ├── 10x_pbmc_5k_V3_figure_generation.r  
│   │   ├── 10x_pbmc_5prime_5k.r  
│   │   ├── 10x_pbmc_5prime_5k_figure_generation.r  
│   │   ├── 10x_pbmc_NextGem.r  
│   │   ├── 10x_pbmc_NextGem_figure_generation.r  
│   ├── dsb_normalize_cluster_pipeline  
│   │   ├── 1_dsb_normalize.r  
│   │   ├── 2_run_umap.r  
│   │   ├── 3_figure_generation.r  
│   │   ├── 4_manual_gate_plots.r  
│   │   ├── generated_data  
│   │   │   ├── MU1Denoised_dsb_Mtx.rds  
│   │   │   ├── NonDenoised_dsb_Mtx.rds  
│   │   │   ├── dsb2_normlog10_list_bybatch.rds  
│   │   │   ├── h1_d0_singlets_ann_Seurat2.4_dsbnorm.rds  
│   │   │   └── h1_sng_metadata_umapdim_prots_dataframe.rds  
│   │   └── manual_gates  
│   │       ├── clr_gates.r  
│   │       └── dsb_gates.r  
│   ├── dsb_process_plots  
│   │   ├── 6_mean_isotype_v_mean_control.R  
│   │   ├── 6a_isotype_figure_generation.r  
│   │   ├── 7_neg_control_plots.R  
│   │   ├── 8_mixture_fits.r  
│   │   └── generated_data  
│   │       ├── b1_dsb_norm_adt_mtx.rds  
│   │       ├── b2_dsb_norm_adt_mtx.rds  
│   │       ├── dsb_norm_adt_mtx.rds  
│   │       ├── isotype_values_dsb.rds  
│   │       ├── logstats.rds  
│   │       ├── mixture_model_metadata_merged.rds  
│   │       ├── multi_component_model_comparison.rds  
│   │       └── normadt.rds   
│   ├── functions  
│   │   ├── analysis_functions.R  
│   │   ├── geneplot4.R  
│   │   └── protein_annotation_functions.r  
│   └── parameter_sensitivity  
│       ├── empty_drop_threshold_batch.r  
│       └── generated_data  
├── data  
│   ├── 10x_data  
│   │   ├── 10x_pbmc10k_V3  
│   │   │   ├── pbmc_10k_protein_v3 — Cell Ranger.htm  
│   │   │   └── raw_feature_bc_matrix  
│   │   │       ├── barcodes.tsv.gz  
│   │   │       ├── features.tsv.gz  
│   │   │       └── matrix.mtx.gz  
│   │   ├── 10x_pbmc5k_NextGem  
│   │   │   ├── 5k_pbmc_protein_v3_nextgem — Cell Ranger.htm  
│   │   │   └── raw_feature_bc_matrix  
│   │   │       ├── barcodes.tsv.gz  
│   │   │       ├── features.tsv.gz  
│   │   │       └── matrix.mtx.gz  
│   │   ├── 10x_pbmc5k_V3  
│   │   │   ├── 5k_pbmc_protein_v3 - 5k Peripheral blood mononuclear cells (PBMCs) from a healthy donor.html  
│   │   │   ├── 5k_pbmc_protein_v3_molecule_info.h5  
│   │   │   └── raw_feature_bc_matrix  
│   │   │       ├── barcodes.tsv.gz  
│   │   │       ├── features.tsv.gz  
│   │   │       └── matrix.mtx.gz  
│   │   └── 10x_pbmc_5prime_5k  
│   │       ├── raw_feature_bc_matrix  
│   │       │   ├── barcodes.tsv.gz  
│   │       │   ├── features.tsv.gz  
│   │       │   └── matrix.mtx.gz  
│   │       └── vdj_v1_hs_pbmc2_5gex_protein — Cell Ranger.htm  
│   ├── V2_Data  
│   │   ├── CITEseq_raw_PMID32094927_seurat2.4.rds  
│   │   ├── H1_day0_scranNorm_adtbatchNorm_dist_clustered_TSNE_labels.rds  
│   │   ├── background_data  
│   │   │   ├── adt_neg_dmx.rds  
│   │   │   ├── adt_neg_dmx_list.rds  
│   │   │   ├── adt_neg_full_list.rds  
│   │   │   └── adt_neg_full_qc.rds  
├── dsb_normalization.Rproj  

### Project options
make rmd readme used: 
usethis::use_readme_rmd()

### notes 

A review of this code has been conducted, no critical errors exist, and to the best of the authors knowledge, there are no problematic file paths, no local system configuration details, and no passwords or keys included in this code. For questions about the dsb software package, please open an issue at the [dsb github repository](https://github.com/niaid/dsb). 

Primary author(s): Matt Mulè  
Organizational contact information: General: john.tsang AT nih.gov, code: mulemp AT nih.gov [permanent address mattmule AT gmail]
Date of release: Oct 7 2020  
Version: NA  
License details: NA  
Description: code to reproduce analysis of manuscript  
Usage instructions: Provided in this markdown  
Example(s) of usage: NA  
Proper attribution to others, when applicable: NA

### code check 

```{r, eval = FALSE}

# checked repo for PII and searched for strings with paths 

# code check 
library(lintr)
fcn = suppressMessages(list.files(here("functions"), ".r", full.names = TRUE))
tenx = suppressMessages(list.files(here("V2/10x_analysis/"),".r", full.names = TRUE))
pipel = suppressMessages(list.files(here("V2/dsb_normalize_cluster_pipeline/"),pattern = ".r", full.names = TRUE))
process = suppressMessages(list.files(here("V2/dsb_process_plots/"),pattern = ".r", full.names = TRUE))
param = suppressMessages(list.files(here("V2/parameter_sensitivity/"),pattern = ".r", full.names = TRUE))

# code check
scp = c(fcn, tenx, pipel, process, param) %>% as.list()
lt = suppressMessages((  
  lapply(scp, lintr::lint)
  )


# ignoring warnings on stylistic choices that do not impact code: 
# use of = not the <- operator for assignment 
# trailing whitespace esp. after "+" in ggplot commands 
# use of space after commas e.g. df[1, ] not df[1,]
# use of i %in% 1:length() in for loops instead of seq_along
```




